---
title: "Assignment 5"
author: "Giang Vu"
date: "3/31/2021"
output: pdf_document
---

```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
library(dplyr)
library(plyr)
library(tidyr)
library(ggplot2)
library(purrr)
library(tibble)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,out.width="80%")
```


## Part 1.\

#### 2.\

Using the transformation of $U$ in previous part, I defined a function in R to generate $n$ simulated Cauchy random variables below. I also tested the function with $n=10$\

```{r}
#define function cauchy.sim
cauchy.sim <- function(n) {
  #generate u from uniform[0,1]
  u <- runif(n)
  # Function for the inverse transform
  return(ifelse((u<0|u>1), 0, tan(pi*(u-0.5))))
}

#test with n = 10
set.seed(0)
cauchy.sim(10)
```

#### 3.\

Using the cauchy.sim function in previous part, I simulated with $n=1000$. Below is the histogram of this random variable with $f_X(x)$ overlaid on it.\

```{r}
#apply function with n =1000
set.seed(0)
cauchy.draws <- cauchy.sim(1000)

#histogram
hist(cauchy.draws, prob = T,breaks = 10000, xlim = c(-10,10),ylim = c(0,.4))
y <- seq(-10, 10, .1)
lines(y, 1/(pi*(1+y^2)), col = "blue")
```

Here is the histogram of this random variable with the density of the true Cauchy(0,1) overlaid on it. I could see my simulation is very close to the true density generated using built in R function.\

```{r}
#histogram
hist(cauchy.draws, prob = T,breaks = 50)
y <- seq(-500, 500, 1)
lines(y, 1/(pi*(1+y^2)), col = "blue")
lines(y, dcauchy(y,location = 0,scale = 1),col ="red")
```


## Part 2.\

### Problem 2.\

#### 4.\

Function $f(x)$ is defined and a plot for it for values between -3 and 3 is generated.\

```{r}
#define f
f <- function(x){
  return(ifelse((x<(-1) | x>2), 0, (1/9)*(4-x^2)))
}

#plot
x <- seq(-3,3,length=100)
plot(x=x,y=f(x),type = "l",
     ylab = "f(x)")

```

#### 5.\

From the plot, we can see the maximum of $f(x)$ is 0.444, which is achieved when $x=0$. (Or we can take the derivative of $f(x)$ and set it equal to 0 to solve for x, which shows us the maximizer is x = 0).\
We then form the envelope $e(x)$ with $\alpha=1/f.max$ and $g(x)$ as the density for the uniform distribution on [-1,2] as follows.\
$e(x)=g(x)/\alpha=f.max\geq f(x)$\

```{r}
#check max of f(x)
max(f(x))
f.max <- f(0)
f.max  

#define envelope e(x)
e <- function(x) {
  return(ifelse((x < -1 | x > 2), Inf, f.max))
}
```

Here is the plot of the envelope function.\

```{r}
plot(x=x,y=f(x),type = "l",
     ylab = "f(x)")
lines(c(-3, -3), c(0, e(0)), lty = 1)
lines(c(3, 3), c(0, e(1)), lty = 1)
lines(c(3,-3), c(e(0),e(0)), lty = 1)
```

#### 6.\

A program using the Accept-Reject Algorithm is written and the simulated data is saved in vector f.draws

```{r}
n.samps <- 10000   # number of samples desired
n       <- 0		     # counter for number samples accepted
f.draws   <- numeric(n.samps) # initialize the vector of output
set.seed(0)
while (n < n.samps) {
  y <- runif(1,min = -1,max = 2)    #random draw from g
  u <- runif(1)
  if (u < f(y)/e(y)) {
    n        <- n + 1
    f.draws[n] <- y
  }
}

```

#### 7.\

Histogram for simulated data with density f overlaid. The simulated data looks quite close to the density.\

```{r}
#histogram
x <- seq(-3,3,length=100)
hist(f.draws, prob = T, ylab = "f(x)", xlab = "x",
     main = "Histogram of simulated draws")

#add density line
lines(x, f(x), lty = 2)
```

### Problem 3.\

#### 8.\

Function $f(x)$ is defined and a plot for it for values between -5 and 5 is generated.\
I named it f1 to avoid mistaking with function f from Problem 2.\

```{r}
#define f
f1 <- function(x){
  return((1/sqrt(2*pi))*exp((-1/2)*x^2))
}

#plot
x <- seq(-5,5,length=100)
plot(x=x,y=f1(x),type = "l",
     ylab = "f(x)")
```

#### 9.\

Function $e(x)$ is defined and I named it e1 to avoid mistaking it with e from Problem 2.\

```{r}
#define e
e1 <- function(x, alpha){
  stopifnot(length(alpha)==1, (0 < alpha & alpha < 1))
  return((1/(pi*(1+x^2)))/alpha)
}
```


#### 10.\

After playing around and plotting for different values for $\alpha$, I chose my good value to be 0.6. As seen from the plot on [-10,10] below, this envelope is very close to $f(x)$ is always right above $f(x)$ at every value of $x$.

```{r}
x <- seq(-10,10,length=100)
plot(x=x,y=f1(x),type = "l",
     ylab = "f(x)")
lines(x,e1(x,0.6), lty = 2)

legend(5.5,0.4,legend=c("f(x)","e(x)"),lty=1:2,cex=0.8)
```

#### 11.\

A function using the Accept-Reject Algorithm is written and it also takes advantage of the Cauchy simulation function we defined in Part 1.\

```{r}
#define function
normal.sim <- function(n){
  i <- 0 # counter for number samples accepted
  norm.draws <- numeric(n) # initialize the vector of output
  set.seed(0)
  while(i < n) {
    y <- cauchy.sim(1) #draw from g(x, which is cauchy using function in part 1
    u <- runif(1)
    if (u < f1(y)/e1(y,0.6)) {
      i <- i + 1
      norm.draws[i] <- y
    }
  }
  return(norm.draws)
}

#test with n = 10
normal.sim(10)
```

#### 12.\

Applying the function above for $n=10000$ I obtained 10,000 draws from standard normal distribution, and made a histogram with $f(x)$ overlaid on the graph. The simulated draw is quite close to the actual density.\

```{r}
#draw 10000
normal.draws <- normal.sim(10000)

#histogram
x <- seq(-10,10,length=100)
hist(normal.draws, prob = T, ylab = "f(x)", xlab = "x",
     main = "Histogram of simulated draws from N(0,1)")

#add density line
lines(x, f1(x), lty = 2)
```

## Part 3.\

#### 13.\

A while() loop is implemented below, and the result is a vector of 13 numbers that satisfy our requirements.

```{r}
x <- 5
set.seed(0)
i <- 0
x.vals <- c()
while(x > 0) {
  r <- runif(1, min=-2, max = 1)
  i <- i + 1
  x.vals[i] <- x
  x <- x + r
}
```

#### 14.\

```{r}
plot(y=x.vals,x=c(1:13),xlab = "Iteration", ylab = "x-values", 
     type = "o", main = "Random walk values versus Iteration number")
```

#### 15.\

The function is defined below.\

```{r}
#define function
random.walk <- function(x.start=5,plot.walk=TRUE){
  num.steps <- 0
  x.vals <- c()
  while(x.start > 0) {
    r <- runif(1, min=-2, max = 1)
    num.steps <- num.steps + 1
    x.vals[num.steps] <- x.start
    x.start <- x.start + r
  }
  if (plot.walk==TRUE){
    plot(y=x.vals,x=c(1:num.steps),xlab = "Iteration", ylab = "x-values", 
         type = "o", main = "Random walk values versus Iteration number")
  }
  return(list(x.vals=x.vals,num.steps=num.steps))
}
```

Test run twice with default values.\

```{r}
random.walk()
```

```{r}
random.walk()
```

Test run twice with 10 and FALSE as input.\

```{r}
random.walk(10,F)
```

```{r}
random.walk(10,F)
```

#### 16.\

By making 10,000 random walks with x = 5, I estimated the mean number of iterations to be 11.25. Essentially on average the random walk we designed carries out about 11 iterations before it terminates.\

```{r}
#loop for 10000 random walks
iters <- numeric(10000)
for (i in 1:10000) {
  li <- random.walk(5,F)
  iters[i] <- li$num.steps #extract iteration number with each walk
}

#mean of iteration number
mean(iters)
```

#### 17.\

The function is modified to add seed setting argument.\

```{r}
#add seed to function
random.walk <- function(x.start=5,plot.walk=TRUE,seed=NULL){
  num.steps <- 0
  x.vals <- c()
  if (!is.null(seed)){set.seed(seed)} #set seed only when seed is specified in argument by user
  while(x.start > 0) {
    r <- runif(1, min=-2, max = 1)
    num.steps <- num.steps + 1
    x.vals[num.steps] <- x.start
    x.start <- x.start + r
  }
  if (plot.walk==TRUE){
    plot(y=x.vals,x=c(1:num.steps),xlab = "Iteration", ylab = "x-values", 
         type = "o", main = "Random walk values versus Iteration number")
  }
  return(list(x.vals=x.vals,num.steps=num.steps))
}
```

Test run with default arguments
```{r}
#test with default
random.walk()

#test with default again
random.walk()

```

Test run with seed and no plot
```{r}
#test with seed 33
random.walk(seed = 33,plot.walk = F)

#test with seed 33 again
random.walk(seed = 33,plot.walk = F)

```

## Part 4.\

#### 18.\

Run the given code.\

```{r}
g <- function(x) {
       return(sqrt(1-x^2))
     }
plot(seq(0,1,.01),g(seq(0,1,.01)),type="l",col="purple")
```

#### 19.\
Take integral of g(x) on [0,1] and we will get the result of $\frac{\pi}{4}$, this is the area under the curve

\newpage

#### 20.\

With $p(x)$ chosen as the pdf of a uniform distribution on [0,1], $p(x)=1$ and thus $g(x)/p(x)=g(x) = \sqrt{1-x^2}$.\

Using Monte Carlo Integration with 100,000 draws from $p(x)$, the integral is estimated to be 0.7852, which is within 1/1000 of the true value (0.7854) calculated using geometric formulas.\

```{r}
#g(x)/p(x) = g(x) because
g.over.p <- function(x) {
  return(sqrt(1-x^2))
}

set.seed(0)
#estimate using MC integration
mean(g.over.p(runif(100000,min = 0,max = 1))) 

#true value
pi/4

set.seed(0)
#difference is within 1/1000
mean(g.over.p(runif(100000,min = 0,max = 1))) - (pi/4)
```
